`forestecology` package for modeling interspecies competition between trees

### 2021-01-25 README.Rmd
Refactored README upto run_cv()

### Preprocess census data

We first combine data from two repeat censuses into a single `tibble` and compute the growth of all surviving individuals. Note that per Allen and Kim (2020) we first remove resprouts from the second census.

```{r}
data(census_df1_ex, census_df2_ex)

# Filter out resprouts
census_df2_ex_no_r <- census_df2_ex %>%
  filter(!str_detect(codes, "R"))

# Merge both censuses and compute growth. Note we require any species variable
# to be coded as a factor
census_combined_ex <-
  compute_growth(census_df1_ex, census_df2_ex_no_r, id = "ID") %>%
  mutate(sp = to_any_case(sp) %>% as.factor())
```

@@

Our growth model assumes that two individual trees compete if they are less than a pre-specified distance `max_dist` apart. Furthermore, we define a buffer region of size `max_dist` from the boundary of the study region. All trees outside this buffer region will be our "focal" trees of interest since we have complete competitor information on them. All trees inside this buffer region will only be considered as "competitor" trees to "focal" trees.

```{r}
data("ex_study_region")

# Set competitor distance
max_dist <- 1

# Create sf representation of buffer region
buffer_region <- ex_study_region %>% 
  compute_buffer_region(direction = "in", size = max_dist)

# Add buffer variable
census_combined_ex <- census_combined_ex %>%
  add_buffer_variable(direction = "in", size = max_dist, region = ex_study_region)

base_plot <- ggplot() +
  geom_sf(data = ex_study_region, fill = "transparent") +
  geom_sf(data = buffer_region, fill = "transparent", linetype = "dashed")

base_plot + 
  geom_sf(data = census_combined_ex, aes(col = buffer), size = 2)
```

We then manually define the blocks/folds for our spatial cross-validation scheme and convert them to an `sf` object using `sfheaders::sf_polygon()`:

@@

Next we assign each tree to blocks/folds using the `spatialBlock()` function from the [`blockCV`](https://github.com/rvalavi/blockCV) package.

```{r}
ex_cv_grid <- spatialBlock(
  speciesData = census_combined_ex, k = 2, selection = "systematic", blocks = blocks,
  showBlocks = FALSE, verbose = FALSE
)

# Add fold information to data
census_combined_ex <- census_combined_ex %>%
  mutate(foldID = ex_cv_grid$foldID %>% factor())

base_plot + 
  geom_sf(data = census_combined_ex, aes(col = buffer, shape = foldID), size = 2) +
  geom_sf(data = blocks, fill = "transparent", col = "orange")
```

@@

```{r}
ex_cv_grid_sf <- ex_cv_grid$blocks %>%
  st_as_sf()

ggplot() +
  geom_sf(data = ex_cv_grid_sf)
```

### Define focal versus competitor trees

We then create a list-column data frame (via `tidyr::nest()`) where each row represents a focal tree of interest and the `comp` variable contains nested information on each of it's competitors

which has a row for each competing pair of individuals (each pair of individuals within `max_dist` of one another):

```{r}
focal_vs_comp_ex <- census_combined_ex %>%
  create_focal_vs_comp(max_dist, cv_grid_sf = ex_cv_grid_sf, id = "ID")

# Nested view
focal_vs_comp_ex

# Unnested view:
focal_vs_comp_ex %>% 
  unnest(cols = "comp")
```

@@

We then fit our competitor growth model as specified in Allen and Kim (2020)

```{r}
comp_bayes_lm_ex <- focal_vs_comp_ex %>%
  comp_bayes_lm(prior_param = NULL, run_shuffle = FALSE)
```

Since the resulting output is an S3 object of class `comp_bayes_lm`, we can both print and visualize its output: the posterior distribution of all the linear regression parameters.

```{r}
# Object summary
comp_bayes_lm_ex

@@

Furthermore, we append fitted/predicted growth values for each tree using `predict.comp_bayes_lm()` to our data frame and then compute the root mean squared error (RMSE).

```{r}
focal_vs_comp_ex <- focal_vs_comp_ex %>%
  mutate(growth_hat = predict(comp_bayes_lm_ex, focal_vs_comp_ex))

# Compute RMSE
focal_vs_comp_ex %>%
  yardstick::rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
```

@@

Here we repeat the process but with running spatial cross-validation: we fit the data to one fold and apply it to the other. Note the increase in RMSE, reflecting the fact that our original estimate of model error was overly optimistic as it did not account for spatial autocorrelation.

```{r}
ex_bw <- focal_vs_comp_ex %>%
  run_cv(max_dist = max_dist, cv_grid = ex_cv_grid_sf) %>%
  right_join(census_combined_ex, by = c("focal_ID" = "ID"))

# Compute RMSE of spatially cross-validated model
ex_bw %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
```

>>>>

### Compute growth of trees based on census data

We first combine data from two repeat censuses into a single `growth` data frame that has the average annual growth of all trees alive at both censuses that aren't resprouts at the second census per Allen and Kim (2020).

```{r}
growth_ex <-
  compute_growth(
    census_1 = census_1_ex,
    census_2 = census_2_ex %>% filter(!str_detect(codes, "R")),
    id = "ID"
    ) %>%
  mutate(sp = to_any_case(sp) %>% factor())
```

@@

Our growth model assumes that two individual trees compete if they are less than a pre-specified distance `comp_dist` apart. Furthermore, we define a buffer region of size `comp_dist` from the boundary of the study region.

```{r}
# Set competitor distance
comp_dist <- 1

# Add buffer variable to growth data frame
growth_ex <- growth_ex %>%
  add_buffer_variable(direction = "in", size = comp_dist, region = study_region_ex)

# Optional: Create sf representation of buffer region
buffer_region <- study_region_ex %>% 
  compute_buffer_region(direction = "in", size = comp_dist)
```

In the visualization below, the solid line represents the boundary of the study region while the dashed line delimits the buffer region within. All trees outside this buffer region (in red) will be our "focal" trees of interest in our model since we have complete competitor information on all of them. All trees inside this buffer region (in blue) will only be considered as "competitor" trees to "focal" trees.

```{r}
base_plot <- ggplot() +
  geom_sf(data = study_region_ex, fill = "transparent") +
  geom_sf(data = buffer_region, fill = "transparent", linetype = "dashed")

base_plot + 
  geom_sf(data = growth_ex, aes(col = buffer), size = 2)
```

Next we add information pertaining to our spatial cross-validation scheme. We first manually define the spatial blocks that will act as our cross-validation folds and convert them to an `sf` object using `sf_polygon()` function from the `sfheaders` package.

@@

Next we assign each tree to the correct folds using the `foldID` variable of the output returned by the `spatialBlock()` function from the [`blockCV`](https://github.com/rvalavi/blockCV) package.

```{r}
SpatialBlock_ex <- spatialBlock(
  speciesData = growth_ex, k = 2, selection = "systematic", blocks = blocks,
  showBlocks = FALSE, verbose = FALSE
)

growth_ex <- growth_ex %>%
  mutate(foldID = SpatialBlock_ex$foldID %>% factor())
```

@@

```{r}
cv_grid_sf_ex <- SpatialBlock_ex$blocks %>%
  st_as_sf()
ggplot() +
  geom_sf(data = cv_grid_sf_ex)
```

In the visualization below, the spatial blocks that act as our cross-validation folds are delineated in orange. Correspondingly the shape of each point indicates which fold each tree has been assigned to.

```{r}
base_plot + 
  geom_sf(data = growth_ex, aes(col = buffer, shape = foldID), size = 2) +
  geom_sf(data = blocks, fill = "transparent", col = "orange")
```

### Compute focal versus competitor tree information

Based on our `growth` data frame, we now explicitly define all "focal" trees and their respective "competitor" trees in a `focal_vs_comp` data frame. This data frame has rows corresponding to each focal tree and all information about its competitors are saved in list-column variable `comp`. We implemented this nested format using `tidyr::nest()` in order to minimize redundancy, given that the same tree can act as a competitor multiple times.

```{r}
focal_vs_comp_ex <- growth_ex %>%
  create_focal_vs_comp(comp_dist, cv_grid_sf = cv_grid_sf_ex, id = "ID")
focal_vs_comp_ex
```

Using `unnest()` we can fully expand the competitor information saved in the `focal_vs_comp` data frame. For example the tree with `focal_ID` equal to 2 located at (1.5, 2.5) has two competitors within `comp_dist` distance from it.

```{r}
focal_vs_comp_ex %>% 
  unnest(cols = "comp")
```

@@

We then fit our competitor growth model as specified in Allen and Kim (2020).

```{r}
comp_bayes_lm_ex <- focal_vs_comp_ex %>%
  comp_bayes_lm(prior_param = NULL)
```

Since the resulting output is an S3 object of class `comp_bayes_lm` we defined, we can both print and visualize its output: the posterior distribution of all linear regression parameters: the intercept, the slope for dbh for each species, and a matrix of all species pairs competitive effects on growth.

```{r}
# Print
comp_bayes_lm_ex

@@

Furthermore, we can apply a generic `predict()` function to the resulting `comp_bayes_lm` object to obtain fitted/predicted values of this model. We append these `growth_hat` values to our `focal_vs_comp` data frame.

```{r}
focal_vs_comp_ex <- focal_vs_comp_ex %>%
  mutate(growth_hat = predict(comp_bayes_lm_ex, focal_vs_comp_ex))
focal_vs_comp_ex
```
We then compute the root mean squared error (RMSE) of the observed versus fitted growths as a measure of our model's fit.

```{r}
focal_vs_comp_ex %>%
  yardstick::rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
```

@@

Whereas in our example above we fit our model to the entirety of the data and then generated fitted/predicted growths on this same data, we now apply the same model with spatial cross-validation. All the trees in a given fold will be given a turn as the "test" data while the trees is all remaining folds will be the "training" data. We then fit the model to the training data, but compute fitted/predicted growths for the separate and independent data.

```{r}
ex_bw <- focal_vs_comp_ex %>%
  run_cv(comp_dist = comp_dist, cv_grid = cv_grid_sf_ex) %>%
  right_join(growth_ex, by = c("focal_ID" = "ID"))
```

Note the increase in RMSE, reflecting the fact that our original estimate of model error was overly optimistic as it did not account for spatial autocorrelation.

```{r}
ex_bw %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
```

### 2021-01-25 etc/paper/forestecology.Rmd
Refactored README upto run_cv()

```{r}
max_dist <- 7.5
cv_fold_size <- 100
```

The first constant is `max_dist` which defines the maximum distance for a tree's competitive neighborhood. Trees within this distance of each other are assumed to compete while those farther than this distance apart do not. Put differently, all trees within `max_dist` of a focal tree will be considered its competitors (see below). Other studies have estimated the value of `max_dist`; we use an average of estimated values @canham_neighborhood_2004, @uriarte_spatially_2004, @tatsumi2013, @canham_neighborhood_2006.

Furthermore, `max_dist` will define the size of all buffers considered, which will be encoded as a binary variable `buffer` as computed by the `add_buffer_variable()` function. This function takes as input the main `growth_df` data frame, the `size` of the buffer which we set as `max_dist`, and the boundary of the study region encoded as a simple features polygon @pebesma_simple_2018. DESCRIBE SF PACKAGE. In the Big Woods example below we will use a pre-loaded simple features polygon while for the SCBI example we present example code on how to manually construct one.

@@

bw_growth_df <- bw_growth_df %>%
  add_buffer_variable(direction = "in", size = max_dist, 
                      region = study_region_bw)

@@

scbi_growth_df <- scbi_growth_df %>%
  add_buffer_variable(direction = "in", size = max_dist, 
                      region = scbi_study_region)

@@

Next we define `focal_vs_comp` data frames which connects each focal tree in the `growth_df` data frames to the trees in its competitive neighborhood range as defined by the `max_dist` constant. So for example, if `growth_df` consisted of two focal trees with two and three neighbors with `max_dist` respectively, `focal_vs_comp` would be a data frame of 5 rows connecting each focal tree to it's competitors. The `create_focal_vs_comp()` function makes this connection taking as inputs the `growth_df` data frame; the `max_dist` constant defining competitive range; `cv_grid_sf`, giving the cross validation grid; and the `id` variable.

@@

```{r bw-focal-vs-comp}
focal_vs_comp_bw <- bw_growth_df %>%
  create_focal_vs_comp(max_dist, cv_grid_sf = bw_cv_grid, id = "treeID")
```

@@

```{r scbi-focal-vs-comp}
focal_vs_comp_scbi <- scbi_growth_df %>%
  create_focal_vs_comp(max_dist, cv_grid_sf = scbi_cv_grid, id = "stemID")
```

@@

```{r bw-spatial-cv}
cv_bw <- focal_vs_comp_bw %>%
  run_cv(max_dist = max_dist, cv_grid = bw_cv_grid) %>%
  right_join(bw_growth_df, by = c("focal_ID" = "treeID"))

@@

```{r scbi-spatial-cv}
cv_scbi <- focal_vs_comp_scbi %>%
  run_cv(max_dist = max_dist, cv_grid = scbi_cv_grid) %>%
  right_join(scbi_growth_df, by = c("focal_ID" = "treeID"))

@@

```{r bw-posterior-viz-0}
focal_vs_comp_bw <- bw_growth_df %>%
  mutate(sp = family) %>%
  create_focal_vs_comp(max_dist = max_dist, cv_grid_sf = bw_cv_grid, 
                       id = "treeID")

>>>>

```{r}
comp_dist <- 7.5
cv_fold_size <- 100
```

The first constant is `comp_dist` which defines the maximum distance for a tree's competitive neighborhood. Trees within this distance of each other are assumed to compete while those farther than this distance apart do not. Put differently, all trees within `comp_dist` of a focal tree will be considered its competitors (see below). Other studies have estimated the value of `comp_dist`; we use an average of estimated values @canham_neighborhood_2004, @uriarte_spatially_2004, @tatsumi2013, @canham_neighborhood_2006.

Furthermore, `comp_dist` will define the size of all buffers considered, which will be encoded as a binary variable `buffer` as computed by the `add_buffer_variable()` function. This function takes as input the main `growth_df` data frame, the `size` of the buffer which we set as `comp_dist`, and the boundary of the study region encoded as a simple features polygon @pebesma_simple_2018. DESCRIBE SF PACKAGE. In the Big Woods example below we will use a pre-loaded simple features polygon while for the SCBI example we present example code on how to manually construct one.

@@

bw_growth_df <- bw_growth_df %>%
  add_buffer_variable(direction = "in", size = comp_dist, 
                      region = study_region_bw)

@@

scbi_growth_df <- scbi_growth_df %>%
  add_buffer_variable(direction = "in", size = comp_dist, 
                      region = scbi_study_region)

@@

Next we define `focal_vs_comp` data frames which connects each focal tree in the `growth_df` data frames to the trees in its competitive neighborhood range as defined by the `comp_dist` constant. So for example, if `growth_df` consisted of two focal trees with two and three neighbors with `comp_dist` respectively, `focal_vs_comp` would be a data frame of 5 rows connecting each focal tree to it's competitors. The `create_focal_vs_comp()` function makes this connection taking as inputs the `growth_df` data frame; the `comp_dist` constant defining competitive range; `cv_grid_sf`, giving the cross validation grid; and the `id` variable.

@@

```{r bw-focal-vs-comp}
focal_vs_comp_bw <- bw_growth_df %>%
  create_focal_vs_comp(comp_dist, cv_grid_sf = bw_cv_grid, id = "treeID")
```

@@

```{r scbi-focal-vs-comp}
focal_vs_comp_scbi <- scbi_growth_df %>%
  create_focal_vs_comp(comp_dist, cv_grid_sf = scbi_cv_grid, id = "stemID")
```

@@

```{r bw-spatial-cv}
cv_bw <- focal_vs_comp_bw %>%
  run_cv(comp_dist = comp_dist, cv_grid = bw_cv_grid) %>%
  right_join(bw_growth_df, by = c("focal_ID" = "treeID"))

@@

```{r scbi-spatial-cv}
cv_scbi <- focal_vs_comp_scbi %>%
  run_cv(comp_dist = comp_dist, cv_grid = scbi_cv_grid) %>%
  right_join(scbi_growth_df, by = c("focal_ID" = "treeID"))

@@

```{r bw-posterior-viz-0}
focal_vs_comp_bw <- bw_growth_df %>%
  mutate(sp = family) %>%
  create_focal_vs_comp(comp_dist = comp_dist, cv_grid_sf = bw_cv_grid, 
                       id = "treeID")
